---
title: "Estimating a flood inundation probability map"
format: html
author: "Gab Chen, Rachel Zhang, University of Pennsylvania"
date: "3/9/2026"
---

## Introduction

Train and validate a spatial predictive model using data from Calgary. Deploy model to predict a flood in City X. Write a report to the "Director of Disaster Planning" with your findings. To be clear, to do this analysis, you will copy the code from the land conservation markdown (predicting the probability of land conservation) and create new independent variables for a model that predicts probability of flood inundation.

---

## Basie Procedure

1.  Choose City X. Choose a city with similar hydrology, topography, and climate, with available data comparable to Calgary's. Don't choose a city on the ocean, or a city in a desert. 

[The city we choose is: Salt Lake City]

Why Calgary and Salt Lake City?

[Calgary and Salt Lake City share several environmental characteristics that make them suitable for comparative flood modeling. Both cities are located inland in western North America and are strongly shaped by mountain hydrology. Calgary lies near the eastern foothills of the Rocky Mountains and is influenced by the Bow and Elbow river systems, while Salt Lake City sits along the Wasatch Range and is influenced by watersheds draining into the Great Salt Lake basin. In both cases, snowpack accumulation in nearby mountains contributes significantly to seasonal runoff, which can increase flood risk during periods of rapid snowmelt or intense rainfall.]

```{r setup, include=FALSE, message=FALSE, warning=FALSE}

# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,      # show code in output
  warning = FALSE,  # hide warnings
  message = FALSE   # hide package messages
)

# Avoid scientific notation
options(scipen = 999)

# Load libraries
library(caret)
library(pscl)
library(plotROC)
library(pROC)
library(sf)
library(tidyverse)
library(knitr)
library(kableExtra)
library(tigris)
library(viridis)

# Set WD
setwd("C:/Users/Gab/OneDrive/Documents/SPRING 2026/CPLN6750_Land Use and Environmental Modeling/flood_inundation_probability/Midterm")

```

2. Gather open data from Calgary’s open data site (https://data.calgary.ca/Links to an external site.) and City X's open data site as well as other internet sources. Build a fishnet grid for each city with the same sized cells. Calgary flood extents and other data are hosted on canvas in Module 5, they are called midTermProject_Data.zip

```{r load_data, warning = FALSE, message = FALSE, results = "hide"}


```

3. Using what we’ve learned about feature engineering, build as many useful variables describing the natural, hydrological and built environment factors that might help explain flood inundation. Form a hypothesis to motivate your search for data. You must include at least one feature from the watershed analysis. You need to build the exact same features for Calgary AND City X. We will have a feature engineering workshop in class.

[The variables we are choosing are: ]

4. Join your features to the vector Fishnet.

5. Move your Fishnet datasets into R and set up your model workflow based on our Preservation markdown.

6. Build logistic regressions for Calgary. Make a training set and test your model by predicting on a test set.  Set up a workflow with goodness of fit metrics - a confusion matrix for example. Experiment until you find a model with decent predictive performance.

7. Using the predict function, feed the exact same independent variables from City X into your model object. Output a prediction for each cell in that city.

8. Visualize and communicate your results.

---

## Deliverables

A technical report - maximum of 4 pages - in pdf form (or knitted R Markdown document html) due on Canvas.

The audience for your report is the Director. Assume they are technically savvy - they were the ones who asked you to do regression modeling for this task! Use clear language, good organization and simple interpretation.

Your report should contain:

1. An introduction framing the planning motivation for your algorithm and presenting the top-line results. How does this modeling approach work? Why are we doing it this way? What was the result of the analysis?
2. A section outlining your feature selection. This should contain maps or plots related to four of your model features. Annotate with some commentary about why these are useful or un-useful features. This must include at least one watershed feature (even if it’s not significant).
3. A section on your model. This should contain your final logistic regression model summary table, your ROC curve plot, and a table of confusion metrics (True Positives, False Positives, etc). Please discuss the modeling process, the training and testing process, comment on the accuracy of the model, and make a table that uses simple statements to describe the four model outcomes ("A True Positive is...").
4. Support your modeling section with three maps:
-One showing true positives, true negatives, false negatives and false positives for the training set in Calgary.
-Second, your inundation predictions for Calgary (entire dataset);
-Third, predictions for your comparable city.
5. A conclusion section commenting on the quality of your model, its usefulness in your comparable city, and your opinion about the utility of this modeling approach.

---

## Notes
There are three files as illustrated above in Module 5 on Canvas (midtermdata.zip). There is the satellite image with flood inundation in light blue and cloud cover in red. You will have to reclassify into a raster of 0=no inundation and 1=inundation. The pink polygon is the study area which you will note has a greater area than the raster. You can 1) create a fishnet at the study area extent in ArcGIS with this toolLinks to an external site., or in R with sf::st_make_grid.Links to an external site. You will then have to 2) create an unique id field in the fishnet shapefile and 3) use zonal statistics as table to get the raster values into the fishnet. Finally 4) remove the grid cells to the north for which we have no raster data from the fishnet.  You also have a DEM raster as above. For your comparison city, look to their open data site for Digital Elevation Map (DEM). Otherwise, go to EarthExplorerLinks to an external site. and download Shuttle Radar Topography Mission data (SRTMLinks to an external site.) data.















